---
title: "Cardiovascular Disease Prediction"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(factoextra)
```

## Data reading:

The data required is readily available in the open UCI Machine learning repository https://archive.ics.uci.edu/ml/datasets/heart+disease. But it is split up based on location and is in a custom format with the extension `.data`. We can read it as table and merge it. The column names are available in the file `heart-disease.names`.

### Read the data

```{r}
# Read all processed data
 
cleaveland_data <- read.table("./data/processed.cleveland.data", fileEncoding = "UTF-8", sep = ",")
hungarian_data <- read.table("./data/processed.hungarian.data", fileEncoding = "UTF-8", sep = ",")
switzerland_data <- read.table("./data/processed.switzerland.data", fileEncoding = "UTF-8", sep = ",")
va_data <- read.table("./data/processed.va.data", fileEncoding = "UTF-8", sep = ",")
```

### Print the dimensions of read data:

```{r}
print("Dimensions of individual datasets :")
print(dim(cleaveland_data))
print(dim(hungarian_data))
print(dim(switzerland_data))
print(dim(va_data))
```

### Concatinate the data and assign column names:
```{r}
# Concat all the datasets
tmp1 <- rbind(cleaveland_data, hungarian_data)
tmp2 <- rbind(switzerland_data, va_data)
heart_data <- rbind(tmp1, tmp2)

# Column names from heart-disease.names file
colnames(heart_data) <- c("age","sex","cp","trestbps","chol","fbs","restecg","thalach","exang","oldpeak","slope","ca","thal","goal")
summary(heart_data)

print("Dimensions of combined data :")
print(dim(heart_data))
```

### Remove all unnecessary columns with ? or :

```{r}
heart_data [heart_data == "?"] <- NA
heart_data <- drop_na(heart_data)

# After removal of all the data we are down to 299 rows
dim(heart_data)
```


### See if the data types are okay 
```{r}
# print the data types
print(sapply(heart_data, class))
```

### Fix the data types
```{r}
# Data types are wrong, should update it based on data available from heart-disease.names
# Age should be a number
heart_data$age <- as.numeric(heart_data$age) 

# Sex should be a factor (1 = male; 0 = female)
heart_data$sex <- as.factor(heart_data$sex)

# cp - chest pain should be a factor
# Value 1: typical angina
# Value 2: atypical angina
# Value 3: non-anginal pain
# Value 4: asymptomatic
heart_data$cp <- as.factor(heart_data$cp)

# trestbps - resting blood pressure
heart_data$trestbps <- as.numeric(heart_data$trestbps)

# chol - serum cholestoral in mg/dl
heart_data$chol <- as.numeric(heart_data$chol)

# fbs - If fasting blood sugar > 120 mg/dl,   (1 = true; 0 = false)
heart_data$fbs <- as.factor(heart_data$fbs)

# restecg - resting electrocardiographic results
# Value 0: normal
# Value 1: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV)
# Value 2: showing probable or definite left ventricular hypertrophy by Estes' criteria
heart_data$restecg <- as.factor(heart_data$restecg)

# thalach: maximum heart rate achieved
heart_data$thalach <- as.numeric(heart_data$thalach)

# exang: exercise induced angina (1 = yes; 0 = no)
heart_data$exang <- as.factor(heart_data$exang)

# oldpeak = ST depression induced by exercise relative to rest
heart_data$oldpeak <- as.numeric(heart_data$oldpeak)

# slope: the slope of the peak exercise ST segment
# Value 1: upsloping
# Value 2: flat
# Value 3: downsloping
heart_data$slope <- as.factor(heart_data$slope)

# ca: number of major vessels (0-3) colored by flourosopy
heart_data$ca <- as.numeric(heart_data$ca)

# thal: 3 = normal; 6 = fixed defect; 7 = reversable defect
heart_data$thal <- as.factor(as.integer(heart_data$thal))

# goal: It distinguish presence (values 1,2,3,4) from absence (value 0)
heart_data$goal <- as.factor(heart_data$goal)

print("After manual updates of datatype")
# print the data types
print(sapply(heart_data, class))
summary(heart_data)
```

### Write the data as csv to local machine:
```{r}
write.csv(heart_data,"./data/heart_data.csv", row.names = FALSE)
```

## Data exploration:

Plot a graph on likeliness of people having a cardio-vascular problems with 0 as abscense and (1,2,3,4) having problems.

```{r}
ggplot(heart_data, aes(x=as.factor(goal), fill=as.factor(goal) )) +
  geom_bar() +
  xlab("Likeliness of having problems") +
  ggtitle("Graph of likeliness of having cardio vascular problems problems") +
  coord_flip()
```

Plot a graph on likeliness of people having a cardio-vascular problems with 0 as abscense and 1 as having problems.

```{r}
tmp_data <- cbind(heart_data$goal, ifelse(heart_data$goal!= 0, 1, 0))

ggplot(heart_data, aes(x=as.factor(tmp_data[, 2]), fill=as.factor(tmp_data[, 2]) )) +
  geom_bar() +
  xlab("Likeliness of having problems") +
  ggtitle("Graph of people having and not having problems") +
  coord_flip()
```

Apply kmeans to see how the clusters are formed
```{r}
kmeans.res <- kmeans(heart_data, 4)
plot(heart_data, col=kmeans.res$cluster)
```


